Recursive Solutions to First-Order Model Counting

### 2022-07-30 doc/paper/paper.tex
Some changes

\captionsetup[subfigure]{width=\linewidth}
\usetikzlibrary{fit,positioning,trees,cd}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}
\newcommand{\crefrangeconjunction}{--}
\crefname{algocf}{algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand*{\twoheadrightarrowtail}{\mathrel{\rightarrowtail\kern-1.9ex\twoheadrightarrow}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\IE}{\textsc{IE}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\wwp}{w}
\DeclareMathOperator{\wwn}{\overline{w}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Imm}{Im}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\size}{\sigma}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\gr}{gr}
\DeclareMathOperator{\leftlsquigarrow}{\text{\reflectbox{$\rightsquigarrow$}}}
\SetKw{KwRet}{yield}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{applyAllRules}{applyAllRules}
\SetKwFunction{applyGreedyRules}{applyGreedyRules}
\SetKwFunction{applyGreedyRulesToFormulas}{applyGreedyRulesToFormulas}
\SetKwFunction{emptyy}{empty}
\SetKwFunction{get}{get}
\SetKwFunction{put}{put}
\SetKwFunction{constructDomainMap}{constructDomainMap}
\SetKwFunction{identifyRecursion}{identifyRecursion}
\SetKwFunction{generateMaps}{generateMaps}
\SetKwFunction{mergeFcgs}{mergeFcgs}
\SetKwFunction{updateCache}{updCache}
\newtheorem{assumption}{Assumption}
\newtheorem{conjecture}{Conjecture}
\newtheorem{constraint}{Constraint}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{experiment}{Experiment}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\Crefname{constraint}{Constraint}{Constraints}
\Crefname{experiment}{Experiment}{Experiments}
\Crefname{clause}{Clause}{Clauses}
\Crefname{diagram}{Diagram}{Diagrams}
\crefname{line}{line}{lines}

@@

% 1. What is the problem?
% 2. Why is it interesting and important?
% 1) the latest first-order variable elimination algorithm \citep{DBLP:journals/jair/TaghipourFDB13}
% 2) other approaches to lifted inference  \citep{DBLP:conf/uai/BrazO17,DBLP:conf/nips/JhaGMS10,DBLP:journals/ijar/RiguzziBZCL17}
% 3) Unlike most early algorithms for lifted inference that were adaptations of probabilistic inference algorithms such as AND/OR search \citep{DBLP:conf/aaai/GogateD10} and variable elimination \citep{DBLP:conf/ijcai/BrazAR05,DBLP:conf/aaai/MilchZKHK08}, WFOMC is an approach grounded in logic.

@@

TODO: remove the references to the figure.

The main conceptual difference between \textsc{Crane} and \textsc{ForcLift} is
that we utilise labelled directed graphs instead of circuits. The cycles in
these graphs represent recursive calls. See \cref{fig:after} for an illustration
of an example scenario. Suppose the original formula $\phi$ depends on a domain
of size $n \in \mathbb{N}$. \emph{Generalised domain recursion} (GDR), one of
the new compilation rules, transforms $\phi$ into a different formula $\psi$
that has an additional constant and some new \emph{constraints}. After some
additional transformations, the constraints in $\psi$ become `uniform' and can
be removed, replacing the domain of size $n$ with a new domain of size
$n-1$---this is the responsibility of the \emph{constraint removal} (CR)
compilation rule. Afterwards, another new compilation rule recognizes that the
resulting formula $\phi(n-1)$ is just like the input formula $\phi(n)$ except it
refers to a different domain. This observation allows us to add a cycle-forming
edge to the graph (highlighted in blue in \cref{fig:after}), which can be
interpreted as function $f$ relying on $f(n-1)$ to compute $f(n)$. To construct
such graphs, we also introduce a novel search algorithm that replaces the greedy
search used by \textsc{ForcLift}.

@@

\Cref{sec:recprelims} defines the representation used for sentences in FOL as
well as discuss caching and some notational conventions. At the beginning of
\cref{sec:methods}, we formally define the graphs that replace circuits in
representing a solution to a (W)FOMC problem and discuss some related notions.
\Cref{sec:rules} introduces the new compilation rules. \Cref{sec:search} then
describes our new search algorithm that makes fewer assumptions than
\textsc{ForcLift} about the right order in which compilation rules should be
applied. In \cref{sec:interpret}, we discuss how a graph can be interpreted as a
collection of (potentially recursive) functions. Finally, in \cref{sec:results}
we compare \textsc{ForcLift} and \textsc{Crane} on a range of function-counting
problems. We show that \textsc{Crane} performs as well as \textsc{ForcLift} on
the instances that were already solvable by \textsc{ForcLift} but is also able
to handle most of the instances that \textsc{ForcLift} fails on.

@@

In this section, we describe our format for FOMC instances, introduce some
notation, and discuss our caching scheme, which is used to identify
possibilities for a recursive call. Note that although the focus of this paper
is on unweighted model counting, \textsc{ForcLift}'s
\citep{DBLP:conf/ijcai/BroeckTMDR11} support for weights trivially transfers to
\textsc{Crane} as well.

@@

% , and \cref{sec:search} outlines the hybrid search algorithm
% that replaces the greedy search used by \textsc{ForcLift}
% \citep{DBLP:conf/ijcai/BroeckTMDR11}.

% TODO: update (and other references to sections, i.e., the beginning of each
% section and the end of the introduction)
We begin this section by formally defining the graphs that \textsc{Crane} uses
as a generalisation of circuits. Then, \cref{sec:rules} describes three new
compilation rules.

@@

% From circuits to graphs.
A \emph{first-order deterministic decomposable negation normal form
  computational graph} (FCG) is a (weakly connected) directed graph with a
single source, node labels, and ordered outgoing edges.\footnote{Note that
  imposing an ordering on outgoing edges is just a limited version of edge
  labelling.} We denote an FCG as $G = (V, s, N^+, \tau)$, where $V$ is the set
of nodes, and $s \in V$ is the unique source. Function $N^+$ maps each node in
$V$ to a \emph{list} of its direct successors. Node labels consist of two parts:
the \emph{type} and the \emph{parameters}. To avoid clutter, we leave the
parameters implicit and let $\tau$ denote the node-labelling function that maps
each node in $V$ to its type. For each node $v \in V$, the length of list
$N^+(v)$ (i.e., the out-degree of $v$) is determined by its type $\tau(v)$. Most
of the types are as in previous work
\citep{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11}. The type for
non-tree-like edges (denoted by $\Reff$), while used before, is extended to
contain the information necessary to support recursive calls. We also add three
new types:
\begin{itemize}
  \item a type for constraint removal denoted by $\CR$,
  \item a type for generalised domain recursion denoted by $\GDR$ (both with
        out-degree one),
  \item and $\star$---a placeholder type (with out-degree zero) for nodes that
        are going to be replaced.
\end{itemize}
When drawing an FCG, we order outgoing edges from left to right, write node
labels directly on the nodes, and omit irrelevant labels and/or parameters. See
\cref{fig:examplefcg} for an example FCG\@. Its source node has out-degree 1
(i.e., $|N^+(s)| = 1$), label
$\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$, and type $\GDR$ (i.e.,
$\tau(s) = \GDR$).

%\paragraph*{Notation.}
Similarly to \citet{DBLP:conf/ijcai/BroeckTMDR11}, we write $T_p$ for an FCG
that has a node with label $T_p$ (i.e., type $T$ and parameter(s) $p$) and
$\star$'s as all of its direct successors. In particular, as an FCG, $\star$
denotes
$(\{\, s \,\}, s, \{\, s \mapsto \langle\rangle \,\}, \{\, s \mapsto \star \,\})$,
i.e., an FCG with just one node of type $\star$ and no edges. We write $T_p(v)$
for an FCG with one edge from a node labelled $T_{p}$ to some other node $v$
(and no other nodes or edges).

\begin{figure}[t]
  \centering
  \begin{subfigure}{0.49\linewidth}
    \centering
    \begin{tikzpicture}[edge from parent/.style={draw,-Latex}]
      \node[draw,circle,label={[text=blue]1}] (top) {}
      child {
        node[draw,circle,label={[text=blue,xshift=-1mm]2}] {}
        child {node[draw,ellipse] (ref) {$\Reff$}}
        child {node[draw,circle,label={[text=blue,xshift=1mm,yshift=-1mm]3}] {$\star$}}
      }
      child {node[draw,circle,label={[text=blue,xshift=1mm,yshift=-1mm]4}] {$\star$}};
      \draw[-Latex] (ref) to [bend left=45] (top);
    \end{tikzpicture}
    \caption{An example FCG}\label{fig:smallexample1}
  \end{subfigure}
  \begin{subfigure}{0.49\linewidth}
    \centering
    \begin{tikzpicture}[edge from parent/.style={draw,-Latex}]
      \node[draw,circle,label={[text=blue]1}] (top) {}
      child {
        node[draw,circle,label={[text=blue,xshift=-1mm]2}] {}
        child {node[draw,circle,label={[text=blue,xshift=2mm,yshift=-1mm]3}] {$\star$}}
      }
      child {node[draw,circle,label={[text=blue,xshift=1mm,yshift=-1mm]4}] {$\star$}};
    \end{tikzpicture}
    \caption{The underlying tree of
      \cref{fig:smallexample1}}\label{fig:smallexample2}
  \end{subfigure}
  \caption{An FCG and its underlying tree. The integers in blue denote the
    pre-order traversal of the underlying tree.}\label{fig:ordering}
\end{figure}

%\paragraph*{Connecting FCGs with formulas.}
Finally, we introduce a structure that represents a solution to a (W)FOMC
problem while it is still being built. A \emph{chip} is a pair $(G, L)$, where
$G$ is an FCG, and $L$ is a list of formulas, such that $|L|$ is equal to the
number of $\star$'s in $G$. The intuition behind this definition is that $L$
contains formulas that still need to be compiled. Once a formula is compiled, it
replaced one of the $\star$'s in $G$. We say that an FCG is \emph{complete}
(i.e., it represents a \emph{complete solution}) if it has no $\star$'s.
Similarly, a chip is complete if its FCG is complete (or, equivalently, the list
of formulas is empty). Let the \emph{underlying tree} of $G$ be the induced
subgraph of $G$ that omits all $\Reff$ nodes.\footnote{Subsequently presented
  algorithms ensure that the underlying tree is guaranteed to be a tree.} Then
we can define an implicit bijection between the formulas in $L$ and the
$\star$'s in $G$ according to the order in which elements of $L$ are listed and
the pre-order traversal of the underlying tree of $G$. For example, if $G$ is as
in \cref{fig:smallexample1} (with its underlying tree in
\cref{fig:smallexample2}), then $|L| = 2$. Moreover, the first element of $L$ is
associated with the $\star$ labelled 3, and the second with the one labelled 4.

@@

% TODO: do I still need the definition of a chip? What about tau?

@@

\section{Conclusion and Future Work}

@@

In this paper, we showed how a state-of-the-art (W)FOMC algorithm can be
empowered by generalising domain recursion and adding support for cycles in the
graph that encodes a solution. To construct such graphs, \textsc{Crane}
supplements \textsc{ForcLift} \citep{DBLP:conf/ijcai/BroeckTMDR11} with three
new compilation rules and a hybrid search algorithm. The experiments in
\cref{sec:results} revealed a range of counting problems that are liftable to
\textsc{Crane} but not to any other (W)FOMC algorithm. The common thread across
these newly liftable problems is (partial) injectivity. Thus, we can formulate
the following conjecture.

>>>>

\usetikzlibrary{fit,positioning,trees,cd}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}
\newcommand{\crefrangeconjunction}{--}
\crefname{algocf}{algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand*{\twoheadrightarrowtail}{\mathrel{\rightarrowtail\kern-1.9ex\twoheadrightarrow}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\GDR}{\textsc{GDR}}
\DeclareMathOperator{\IE}{\textsc{IE}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}
\DeclareMathOperator{\wwp}{w}
\DeclareMathOperator{\wwn}{\overline{w}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\Imm}{Im}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\size}{\sigma}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\gr}{gr}
\DeclareMathOperator{\leftlsquigarrow}{\text{\reflectbox{$\rightsquigarrow$}}}
\SetKw{KwRet}{yield}
\SetKwProg{Fn}{Function}{:}{}
\SetKwFunction{applyAllRules}{applyAllRules}
\SetKwFunction{applyGreedyRules}{applyGreedyRules}
\SetKwFunction{applyGreedyRulesToFormulas}{applyGreedyRulesToFormulas}
\SetKwFunction{emptyy}{empty}
\SetKwFunction{get}{get}
\SetKwFunction{put}{put}
\SetKwFunction{constructDomainMap}{constructDomainMap}
\SetKwFunction{identifyRecursion}{identifyRecursion}
\SetKwFunction{generateMaps}{generateMaps}
\SetKwFunction{mergeFcgs}{mergeFcgs}
\SetKwFunction{updateCache}{updCache}
\newtheorem{assumption}{Assumption}
\newtheorem{conjecture}{Conjecture}
\newtheorem{constraint}{Constraint}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{experiment}{Experiment}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\Crefname{constraint}{Constraint}{Constraints}
\Crefname{experiment}{Experiment}{Experiments}
\Crefname{clause}{Clause}{Clauses}
\Crefname{diagram}{Diagram}{Diagrams}
\crefname{line}{line}{lines}

@@

% 1. What is the problem?
% 2. Why is it interesting and important?

@@

The main conceptual difference between \textsc{Crane} and \textsc{ForcLift} is
that we utilise labelled directed graphs instead of circuits. The cycles in
these graphs represent recursive calls. Suppose the input formula $\phi$ depends
on a domain of size $n \in \mathbb{N}$. \emph{Generalised domain recursion}
(GDR), one of the new compilation rules, transforms $\phi$ into a different
formula $\psi$ that has an additional constant and some new \emph{constraints}.
After some additional transformations, the constraints in $\psi$ become
`uniform' and can be removed, replacing the domain of size $n$ with a new domain
of size $n-1$---this is the responsibility of the \emph{constraint removal} (CR)
compilation rule. Afterwards, another new compilation rule recognizes that the
resulting formula $\phi(n-1)$ is just like the input formula $\phi(n)$ except it
refers to a different domain. This observation allows us to add a cycle-forming
edge to the graph, which can be interpreted as function $f$ relying on $f(n-1)$
to compute $f(n)$.

@@

We begin by defining the representation used for sentences in FOL in
\cref{sec:recprelims}. Then, in \cref{sec:methods}, we define the graphs that
replace circuits in representing a solution to a (W)FOMC problem.
\Cref{sec:rules} introduces the new compilation rules. In \cref{sec:interpret},
we discuss how a graph can be interpreted as a collection of (potentially
recursive) functions. In \cref{sec:results}, we compare \textsc{ForcLift} and
\textsc{Crane} on a range of function-counting problems. We show that
\textsc{Crane} performs as well as \textsc{ForcLift} on the instances that were
already solvable by \textsc{ForcLift} but is also able to handle most of the
instances that \textsc{ForcLift} fails on. Finally, in \cref{sec:conclusion}, we
conclude by outlining some conjectures and directions for future work.

@@

@@

@@

% From circuits to graphs.
A \emph{first-order deterministic decomposable negation normal form
  computational graph} (FCG) is a (weakly connected) directed graph with a
single source, node labels, and ordered outgoing edges.\footnote{Note that
  imposing an ordering on outgoing edges is just a limited version of edge
  labelling.} Node labels consist of two parts: the \emph{type} and the
\emph{parameters}. The out-degree of each node is determined by its type.
Similarly to \citet{DBLP:conf/ijcai/BroeckTMDR11}, we write $T_p$ for an FCG
that has a node with label $T_p$ (i.e., type $T$ and parameter(s) $p$) and
$\star$'s as all of its direct successors.

Most of the types are as in previous work
\citep{DBLP:conf/nips/Broeck11,DBLP:conf/ijcai/BroeckTMDR11}. The type for
non-tree-like edges (denoted by $\Reff$), while used before, is extended to
contain the information necessary to support recursive calls. We also add three
new types:
\begin{itemize}
  \item a type for constraint removal denoted by $\CR$,
  \item a type for generalised domain recursion denoted by $\GDR$ (both with
        out-degree one),
  \item and $\star$---a placeholder type (with out-degree zero) for nodes that
        are going to be replaced.
\end{itemize}
When drawing an FCG, we order outgoing edges from left to right, write node
labels directly on the nodes, and omit irrelevant labels and/or parameters. See
\cref{fig:examplefcg} for an example FCG\@. Its source node has out-degree 1,
label $\GDR_{\Gamma \gets \Gamma \setminus \{\, x \,\}}$, and type $\GDR$.

% Connecting FCGs with formulas
Finally, we introduce a structure that represents a solution to a (W)FOMC
problem while it is still being built. A \emph{chip} is a pair $(G, L)$, where
$G$ is an FCG, and $L$ is a list of formulas, such that $|L|$ is equal to the
number of $\star$'s in $G$. The intuition behind this definition is that $L$
contains formulas that still need to be compiled. Once a formula is compiled, it
replaced one of the $\star$'s in $G$ according to a set order. We say that an
FCG is \emph{complete} (i.e., it represents a \emph{complete solution}) if it
has no $\star$'s. Similarly, a chip is complete if its FCG is complete (or,
equivalently, the list of formulas is empty).

@@

@@

\section{Conclusion and Future Work}\label{sec:conclusion}

@@

In this paper, we showed how a state-of-the-art (W)FOMC algorithm can be
empowered by generalising domain recursion and adding support for cycles in the
graph that encodes a solution. To construct such graphs, \textsc{Crane}
supplements \textsc{ForcLift} \citep{DBLP:conf/ijcai/BroeckTMDR11} with three
new compilation rules. The experiments in \cref{sec:results} revealed a range of
counting problems that are liftable to \textsc{Crane} but not to any other
(W)FOMC algorithm. Although here we focused on FOMC problems,
\textsc{ForcLift}'s support for weights trivially transfers to \textsc{Crane} as
well. The common thread across these newly liftable problems is (partial)
injectivity. Thus, we can formulate the following conjecture.
